---
title: Creating a trial-based time variable
author: ~
date: '2017-07-19'
slug: creating-a-trial-based-time-variable
tags:
  - R
  - eye tracking
description: 'Raw eye tracker data does not come with a trial-based timestamp variable. That is, there is no variable that starts at 0 when a trial begins and continues counting until the next trial. In this post I show how to create that variable based on a timestamp and trial information.'
---
```{r setup, include = FALSE}
# Set knitr options
knitr::opts_chunk$set(echo = TRUE)

# Packages
library(tidyverse)
library(knitr)

# Read data
data <- read_tsv("../../static/data/trial-based-time-variable-data.tsv")
```

One of the first issues I ran into when starting to analyze eye tracker data was that the raw data does not contain a trial-based time variable. I expected that the timestamp variable would be a variable that starts at 0 when a trial begins and that it would keep adding up at a rate equal to the sampling frequency, until the end of the trial. Instead, you are likely to get a timestamp variable that looks like a random set of numbers, as shown here:

```{r echo = FALSE, results = "asis"}
# Print data
kable(head(data, 10), caption = "Example data")
```

The **timestamp** variable actually reflects the internal clock of the hardware used to get the data. This means that each measurement recording has a specific clock time associated with it, rather than a time stamp related to an event in the experiment.

Fortunately, it's relatively easy to turn this variable into a more useful variable. What we want is a variable, say **time**, that starts at 0 when a trial begins. The subsequent measures, in the same trial, should then be timed relative to the start of the trial. We want this for every trial.

The required steps to get this variable are as follows:

1. For each trial, get the minimum of the **timestamp** variable (e.g., 212275472 in trial 1)
2. Repeat this value across the entire trial
3. Subtract this value from the **timestamp** variable

In R, this would look like this:

```{r create_time_variable}

# Create a trial-based time variable
data %>%
    group_by(subject, trial) %>%
    mutate(time = timestamp - min(timestamp)) -> data

# Divide by a 1000 to have the variable denote milliseconds rather than microseconds
data$time <- data$time / 1000
```

We take our data frame, group the data by **subject** and **trial** (because we want the minimum for each individual trial), and create a new variable called **time** that is equal to the value in **timestamp** minus the minimum of the timestamp for that trial. Additionally, we divide the result by a 1000 because the internal clock is in microseconds, and I prefer milliseconds. The result is this:


```{r echo = FALSE, results = "asis"}
# Print data
kable(head(data, 10), caption = "Example data with time variable")
```

We see that our new variable time indeed starts at 0, adds up until the next trial, where it starts at 0 again. Excellent!
